---
title: "Modelling techniques in OMPR"
author: "Dirk Schumacher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignettes discribes the modelling techniques available in `ompr` to make your life easier when developing a mixed integer programming model. 

## A MIP Model

You can think of a MIP Model as a big constraint maxtrix and a set of vectors. But you can also think of it as a set of decision variables, an objective function and a number of constraints. `ompr` implements the latter approach.

For example, [Wikipedia](https://en.wikipedia.org/wiki/Knapsack_problem) describes the Knapsack problem like this:
$$
\begin{equation*}
\begin{array}{ll@{}ll}
\text{max}  & \displaystyle\sum\limits_{i=1}^{n} v_{i}x_{i} & &\\
\text{subject to}& \displaystyle\sum\limits_{i=1}^{n}   w_{i}x_{i} \leq W,  & &\\
                 &                                                x_{i} \in \{0,1\}, &i=1 ,\ldots, n&
\end{array}
\end{equation*}
$$

This is the `ompr` equivalent:

```{r, eval=FALSE}
n <- 10; W <- 2
v <- runif(n);w <- runif(n)
model <- MIPModel() %>% 
  add_variable(x[i], i = 1:n, type = "binary") %>% 
  set_objective(sum_expr(v[i] * x[i], i = 1:n)) %>% 
  add_constraint(sum_expr(w[i] * x[i], i = 1:n) <= W)
```


The overall idea is to use modern R idioms to construct models like the one above as readable as possible directly in R. `ompr` will do the heavy lifting and transforms everything into matrices/vectors and pass it to your favorite solver.

```{r, echo=FALSE}
library(ompr)
library(magrittr)
```


## Pipes

Each function in `ompr` creates immutable copies of the models. In addition the function interface has been designed to work with `magrittr` pipes. You always start with an empty model and add components to it.

```{r}
MIPModel() %>% 
  add_variable(x) %>% 
  set_objective(x) %>% 
  add_constraint(x <= 1)
```

## Variable types

Variables can be of type `continuous`, `integer` or `binary`.

```{r}
MIPModel() %>% 
  add_variable(x, type = "integer") %>% 
  add_variable(y, type = "continuous") %>% 
  add_variable(z, type = "binary")
```

## Variable bounds

Variables can have lower and upper bounds. 

```{r}
MIPModel() %>% 
  add_variable(x, lb = 10) %>% 
  add_variable(y, lb = 5, ub = 10)
```

## Indexed variables

Often when you develop a complex model you work with indexed variables. This is an important concept `ompr` supports.

```{r}
MIPModel() %>% 
  add_variable(x[i], i = 1:10) %>%  # creates 10 decision variables
  set_objective(x[5]) %>% 
  add_constraint(x[5] <= 10)
```

## Summation over variables

If you have indexed variables then you often want to sum over a subset of variables. 

The following code creates a model with three decision variables $x_1$, $x_2$, $x_3$. An objective function $\sum_i x_i$ and one constraint $\sum_i x_i \leq 10$.

```{r}
MIPModel() %>% 
  add_variable(x[i], i = 1:3) %>% 
  set_objective(sum_expr(x[i], i = 1:3)) %>% 
  add_constraint(sum_expr(x[i], i = 1:3) <= 10)
```

## Quantifiers

`add_variable`, `add_constraint`, `set_bounds`, `sum_expr` all support a common quantifier interface that also supports filter expression. A more complex example will show what that means.

```{r}
MIPModel() %>% 
  # Create x_{i, j} variables for all combinations of i and j where
  # i = 1:10 and j = 1:10.
  add_variable(x[i, j], type = "binary", i = 1:10, j = 1:10) %>% 
  
  # add a y_i variable for all i between 1 and 10 with i mod 2 = 0
  add_variable(y[i], type = "binary", i = 1:10, i %% 2 == 0) %>% 
  
  # we maximize all x_{i,j} where i = j + 1
  set_objective(sum_expr(x[i, j], i = 1:10, j = 1:10, i == j + 1)) %>% 
  
  # for each i between 1 and 10 with i mod 2 = 0
  # we add a constraint \sum_j x_{i,j}
  add_constraint(sum_expr(x[i, j], j = 1:10) <= 1, i = 1:10, i %% 2 == 0) %>% 
  
  # of course you can leave out filters or add more than 1
  add_constraint(sum_expr(x[i, j], j = 1:10) <= 2, i = 1:10) 
```

## Special bounds on a subset of variables

Imagine you want to model a matching problem with a single binary decision variable $x_{i,j}$ that is $1$ iff object $i$ is matched to object $j$. One constraint would be to allow matches only if $i \neq j$. This can be modelled by a constraint or by selectively changing bounds on variables. The latter approach can be used by solvers to improve the solution process.

```{r}
MIPModel() %>% 
  add_variable(x[i, j], i = 1:10, j = 1:10, 
               type = "integer", lb = 0, ub = 1) %>% 
  set_objective(sum_expr(x[i, j], i = 1:10, j = 1:10)) %>% 
  add_constraint(x[i, i] == 0, i = 1:10) %>% 
  
   # this sets the ub to 0 without adding new constraints
  set_bounds(x[i, i], ub = 0, i = 1:10)
```

## External model parameters

Of course you will need external parameters for your models. You can reuse any variable defined in your R environment within the MIP Model.

```{r}
n <- 5 # number of our variables
costs <- rpois(n, lambda = 3) # a cost vector
max_elements <- 3
MIPModel() %>% 
  add_variable(x[i], type = "binary", i = 1:n) %>% 
  set_objective(sum_expr(costs[i] * x[i], i = 1:n)) %>% 
  add_constraint(sum_expr(x[i], i = 1:n) <= max_elements)
```

## Extract model solutions

Once you have a model, you pass it to a solver and get back a solutions. The main interface to extract variable values from a solution is the function `get_solution`. It returns a data.frame for indexed variable and thus makes it easy to subsequently use the value.

We use `ROI` and `GLPK` to solve it.

```{r}
library(ROI)
library(ROI.plugin.glpk)
library(ompr.roi)
```

```{r}
set.seed(1)
n <- 5
weights <- matrix(rpois(n * n, 5), ncol = n, nrow = n)
result <- MIPModel() %>% 
  add_variable(x[i, j], i = 1:n, j = 1:n, type = "binary") %>% 
  set_objective(sum_expr(weights[i, j] * x[i, j], i = 1:n, j = 1:n)) %>% 
  add_constraint(sum_expr(x[i, j], j = 1:n) == 1, i = 1:n) %>% 
  solve_model(with_ROI("glpk", verbose = TRUE))
```

```{r}
get_solution(result, x[i, j]) %>% 
  dplyr::filter(value == 1)
```

You can also fix certain indexes.

```{r}
get_solution(result, x[2, j])
```


