---
title: "Assigning students to courses"
author: "Dirk Schumacher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assign students to courses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction

In this article we will look at [assignment problems](https://en.wikipedia.org/wiki/Generalized_assignment_problem). 

As a real world example problem we would like to match a group of students to a set of courses with the following constraints:

* Each course has a capacity
* Every student needs to be assigned to exactly one course. 
* All students have stated individual preferences on a scale from 1 to 3, where 3 is the most favorable course.

## The data

We have $n$ students:
```{r}
n <- 40
```

And $m$ courses with equal capacity. The capacity can vary among courses though.
```{r}
m <- 4
capacity <- rep.int(11, m) # all have equal capacities
```

In addition, each student has three preferences. To model this we have a function that gives us three courses for each student. The first component has perference 1, second 2, and third 3:
```{r}
set.seed(1234)
preferences <- memoise::memoise(function(student) sample(seq_len(m), 3))
```

```{r}
preferences(1)
```

The last component we need is a weight functions to make the model formulation easier.
This function gives us the preference weighting for a course and student pair.

```{r}
# the weight of a student choosing a course
# if the course is not among the preferences, the weight is -100000
weight <- function(student, course) {
  p <- which(as.numeric(course) == preferences(as.numeric(student)))
  as.integer(if (length(p) == 0) {
    -100000
  } else {
    p
  })
}
```

Some examples:
```{r}
weight(1, 3)
weight(1, 23) # this was not a choice by student 1, so we give it a big penalty
```

Let's take a look at our random preferences. We plot the number of votes for each course grouped by the preference (1, 2, 3).
```{r, message=FALSE}
library(ggplot2)
library(purrr)
library(dplyr)
plot_data <- expand.grid(
  course = seq_len(m),
  weight = 1:3
) %>% rowwise() %>% 
  mutate(count = sum(map_int(seq_len(n), ~weight(.x, course) == weight))) %>% 
  mutate(course = factor(course))
ggplot(plot_data, aes(x = course, y = count, fill = weight)) + 
  geom_bar(stat = "identity") + 
  viridis::scale_fill_viridis() + 
  geom_hline(yintercept = 11)
```

## The model

The idea is to introduce a binary variable $x_{i, j}$ that is $1$ if student $i$ is matched to course $j$. As an objective we will try to satisfy preferences according to their weight. So assigning a student to a course with preference 3 gives 3 points and so forth. The model assumes, that the total capacity of the courses is enough for all students.

Here it is in mathematical notation:

$$
\begin{equation*}
\begin{array}{ll@{}ll}
\text{max}  & \displaystyle\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{m}weight_{i,j} \cdot x_{i, j} & &\\
\text{subject to}& \displaystyle\sum\limits_{i=1}^{n}   x_{i, j} \leq capacity_j,  & j=1 ,\ldots, m&\\
                 & \displaystyle\sum\limits_{j=1}^{m}   x_{i, j} = 1,  & i=1 ,\ldots, n&\\
                 &                                                x_{i,j} \in \{0,1\}, &i=1 ,\ldots, n, & j=1 ,\ldots, m
\end{array}
\end{equation*}
$$

Or directly in R:

```{r, message=FALSE}
library(ompr)
model <- MIPModel() %>%
  
  # 1 iff student i is assigned to course m
  add_variable(x[i, j], i = 1:n, j = 1:m, type = "binary") %>%
  
  # maximize the preferences
  set_objective(sum_expr(weight(i, j) * x[i, j], i = 1:n, j = 1:m)) %>%
  
  # we cannot exceed the capacity of a course
  add_constraint(sum_expr(x[i, j], i = 1:n) <= capacity[j], j = 1:m) %>% 
  
  # each student needs to be assigned to one course
  add_constraint(sum_expr(x[i, j], j = 1:m) == 1, i = 1:n)
model
```

## Solve the model

We will use `glpk` to solve the above model.

```{r}
library(ompr.roi)
library(ROI.plugin.glpk)
result <- solve_model(model, with_ROI(solver = "glpk", verbose = TRUE))
```

We solved the problem with an objective value of `r objective_value(result)`.

```{r}
matching <- result %>% 
  get_solution(x[i,j]) %>%
  filter(value > .9) %>%  
  select(i, j) %>% 
  rowwise() %>% 
  mutate(weight = weight(as.numeric(i), as.numeric(j)), 
         preferences = paste0(preferences(as.numeric(i)), collapse = ",")) %>% ungroup
```

```{r}
head(matching)
```


```{r}
matching %>% 
  group_by(weight) %>% 
  summarise(count = n())
```



`r nrow(filter(matching, weight == 3))` students got their top preference.
`r nrow(filter(matching, weight == 2))` students were assigned to their second choice and `r nrow(filter(matching, weight == 1))` students got their least preferable course.

The courses have the following number of students:

```{r}
matching %>% 
  group_by(j) %>% 
  summarise(count = n())
```



## Feedback

Do you have any questions, ideas, comments? Or did you find a mistake? Let's discuss on [Github](https://github.com/dirkschumacher/ompr/issues).
