---
title: "The Minimum Graph Coloring Problem"
author: "Dirk Schumacher"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Minimum Graph Coloring Problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

In this article we will model the [minimum graph coloring problem](https://en.wikipedia.org/wiki/Graph_coloring). The goal: color a map with as few colors as possible while no two adjacent regions having the same color.

## Read spatial data

First, let's load some useful packages needed for (spatial) data processing.

```{r, message=FALSE, warning=FALSE}
library(rgeos)
library(rgdal)
library(maptools)
library(dplyr)
```

Then we read in the 50 states of the US.

```{r}
# data from https://github.com/datasets/geo-boundaries-us-110m
# reference data from here: https://github.com/nvkelso/natural-earth-vector/tree/master/110m_cultural
# License of reference data: public domain (https://github.com/nvkelso/natural-earth-vector/blob/master/LICENSE.md)
map_data <- rgdal::readOGR("https://raw.githubusercontent.com/datasets/geo-boundaries-us-110m/84e946f6b1de01e2642bcdb17d5b697acb6b48c4/json/ne_110m_admin_1_states_provinces_shp_scale_rank.geojson", "OGRGeoJSON")
```

Next step is to create an adjacency list to determine neighboring states.

```{r}
# this gives as an adjancy list
neighbors <- spdep::poly2nb(map_data)

# a helper function that determines if two nodes are adjacent
is_adjacent <- function(i, j) {
  purrr::map2_lgl(i, j, ~ .y %in% neighbors[[.x]])
}
```

We can then ask:
```{r}
is_adjacent(1, 2)
is_adjacent(1, 3)
```


## Optimization model

Next, we will model the problem with `ompr` as a mixed integer linear program that tries to find a coloring with as few colors as possible.

```{r, message=FALSE}
library(ompr)
```

```{r, message=FALSE}
n <- nrow(map_data@data) # number of nodes aka states
max_colors <- 4 # 4 should be enough. But you increase this number.

# based on the formulation from here
# http://wwwhome.math.utwente.nl/~uetzm/do/IP-FKS.pdf
model <- MIPModel() %>% 
  
  # 1 iff node i has color k
  add_variable(x[i, k], type = "binary", i = 1:n, k = 1:max_colors) %>% 
  
  # 1 iff color k is used
  add_variable(y[k], type = "binary", k = 1:max_colors) %>% 
  
  # minimize colors
  set_objective(sum_expr(y[k], k = 1:max_colors), direction = "min") %>% 
  
  # each node is colored
  add_constraint(sum_expr(x[i, k], k = 1:max_colors) == 1, i = 1:n) %>% 
  
  # if a color k is used, set y[k] to 1
  add_constraint(sum_expr(x[i, k], i = 1:n) <= n * y[k], k = 1:max_colors) %>% 
  
  # no adjacent nodes have the same color
  add_constraint(x[i, k] + x[j, k] <= 1, i = 1:n, j = 1:n, k = 1:max_colors, is_adjacent(i, j))
```

```{r}
model
```

The last constraint uses *filter expressions*:
```{r, eval=FALSE}
add_constraint(x[i, k] + x[j, k] <= 1, i = 1:n, j = 1:n, k = 1:max_colors, is_adjacent(i, j))
```
This code generates a constraint for each combination of i, j and k with the additional constraint, that `is_adjacent(i, j)` needs to evaluate to `TRUE`. Using *filter expressions* let's you design a very specific set of constraints with minimal code. This principle can be used with all functions that acceppt quantifiers: `sum_expr`, `set_bounds`, `add_constraint` and `add_variable`.

## Solve it

Now since we have built the model, we will solve it using the excellent `ROI` package and the `symphony` solver. `ompr.roi` is a solver that takes an `ompr` package and translates it to `ROI`: this makes it possible to use range of solvers out of the box.

```{r, message=FALSE}
library(ROI)
library(ROI.plugin.symphony)
library(ompr.roi)
```

```{r, warning=FALSE}
result <- solve_model(model, with_ROI("symphony"))
```

```{r}
result
```

Yay. We found the minimal coloring with `r objective_value(result)` colors.

## Plot the result

Last step is to plot the result. First we will get the colors from the optimal solution.

```{r}
assigned_colors <- get_solution(result, x[i, k]) %>% 
  filter(value > 0.9) %>% 
  arrange(i)
```

```{r}
head(assigned_colors, 5)
```

Then we need to prepare the data for ggplot and join the colors to the data.

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
color_data <- map_data@data
color_data$color <- assigned_colors$k
plot_data_fort <- fortify(map_data, region = "adm1_code") %>% 
  left_join(select(color_data, adm1_code, color), 
            by = c("id" = "adm1_code")) %>% 
  mutate(color = factor(color))
```

Now we have everything to plot it:

```{r}
ggplot(plot_data_fort, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = color)) + 
  coord_quickmap() + 
  viridis::scale_fill_viridis(discrete = TRUE, option = "D")
```

